---
title: "ch13 clustering"
author: "Bernie Mulvey"
date: "2023-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height = 10,fig.width = 7,include = FALSE)
knitr::opts_chunk$set(fig.width=7,fig.height=10)
#### sets tab autocompletion for directories to begin from the directory containing the .Rproj session, instead of the script's directory if different
knitr::opts_knit$set(root.dir = here::here())

library(ggplot2)
library(data.table)
library(Biostrings)
library(gridExtra)
require(colorout)
ColorOut()
options("styler.addins_style_transformer" = "biocthis::bioc_style()")
library(SpatialExperiment)
library(ggspavis)
library(scater) # addPerCellQC
library(nnSVG)
library(BiocParallel)
library(scran)
library(parallel)
# library(STexampleData)

theme_set(theme_bw()+theme(axis.text.x = element_text(size = 14), axis.title.x = element_text(size = 16), axis.text.y = element_text(size = 14), axis.title.y = element_text(size =16), plot.title = element_text(size = 20,hjust=0.5), strip.text = element_text(size=18), legend.text = element_text(size=10), legend.title = element_text(size=11,hjust=0.5)))
```

load previous objs
```{r}
# filtered data
splc2 <- readRDS("splc2_line201.RDS")
# hvgs, svgs, and combinations
vgs <- fread("Top 1310 ea SVGs and HVGs, union, and n union top SVGs.txt")

# umap and pca objs
umlist <- readRDS("processed/ch12/umaplist.RDS")
umdirlist <- readRDS("processed/ch12/umapdirlist.RDS")
pcalist <- readRDS("processed/ch12/pcalist.RDS")
```

Spatially-UNAWARE clustering:
"We can perform clustering by applying standard clustering methods developed for single-cell RNA sequencing data, using molecular features (gene expression). Here, we apply graph-based clustering using the Walktrap method implemented in scran (Lun, McCarthy, and Marioni 2016), applied to the top 50 PCs calculated on the set of top HVGs."

Since we have H, S, and H+S vgs, we can try it w all four. the bigger question will be what to set k.

```{r}
bpparam <- MulticoreParam(workers=8)
glist <- bplapply(pcalist,BPPARAM = bpparam,FUN=function(x){
  scran::buildSNNGraph(x,k=16,use.dimred=reducedDimNames(x))
})

g_walk <- bplapply(glist,igraph::cluster_walktrap,BPPARAM = bpparam)
walkclusts <- lapply(g_walk,FUN=function(x){x$membership})

# output is a vector of length(colData((spe)) with the cluster assignment 
walkclusts[[1]]

## check cluster sizes
lapply(walkclusts,table)
## notice that despite k set to 16 above, the different gene sets give 8-14 clusters. not sure what to make of that. 

# append cluster labels to spes for visualization
names(walkclusts) <- names(pcalist)
i<-1
for (i in c(1:4)){
  colLabels(pcalist[[i]]) <- factor(walkclusts[[i]])
}
rm(i,g_walk,glist)
gc(full=T)
# colLabels gives this column the name label and no way to do otherwise
names(colData(pcalist[[1]]))
# what does it do when you feed a second label in?
colLabels(pcalist[[1]]) <- rep("z",ncol(pcalist[[1]]))
colData(pcalist[[1]])[1:3,32]
## oh, just overwrites it. so hang onto walkclusts so we can rotate through them and other labels as we go. and switch the pcal [[1]] back to the walkclust assignments
colLabels(pcalist[[1]]) <- factor(walkclusts[[1]])
# saveRDS(walkclusts,"data/pcalist_walktrap_clusts.RDS")
```

plot cluster assignments onto spatial data
### note: scCustomize has functions to pull from palette sets of up to 50 colors
### https://samuel-marsh.github.io/scCustomize/articles/Color_Palettes.html
### and see here for storing those palette assignments https://samuel-marsh.github.io/scCustomize/articles/Helpers_and_Utilities.html
```{r}
library(scCustomize)

# ick, the first two colors are black and gray so skip those ones by pulling 2 extra colors and only using colors ≥3
plts <- list()
i<-1
for (i in c(1:length(pcalist))){
  x<-pcalist[[i]]
  plts[[i]] <- ggspavis::plotSpots(pcalist[[i]], annotate = "label", palette = DiscretePalette_scCustomize(num_colors = 2 + length(unique(colData(x)$label)), palette = "polychrome")[3:(length(unique(colData(x)$label)) + 2)])+
    ggtitle(paste0(names(pcalist)[i]))
  rm(x)
}
rm(i)

# pdf("plots/ch13_clusters_walktrap_k16_by_hsvgSets.pdf",height=18,width=4)
do.call("grid.arrange",c(plts,ncol=1))
# dev.off()
rm(plts)

```


### now i'm not really sure what this walktrap business is...i think we need to play with other spatially-aware and unaware clustering algos. let's look through the LC and dlPFC git repos for what some more refined clustering techniques might look like.