---
title: "ch13 clustering"
author: "Bernie Mulvey"
date: "2023-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height = 10,fig.width = 7,include = FALSE)
knitr::opts_chunk$set(fig.width=7,fig.height=10)
#### sets tab autocompletion for directories to begin from the directory containing the .Rproj session, instead of the script's directory if different
knitr::opts_knit$set(root.dir = here::here())

library(data.table)
require(colorout)
ColorOut()
options("styler.addins_style_transformer" = "biocthis::bioc_style()")
library(SpatialExperiment)
# library(ggspavis)
# library(scater) # addPerCellQC
# library(nnSVG)
library(BiocParallel)
library(scran)
library(parallel)
# library(scCustomize)
library(fasthplus)
# library(STexampleData)
library(segmented)

```

load only objs we need for mem effish
```{r}
pcalist <- readRDS("processed/ch12/pcalist.RDS")
```

### Extending from ch13, part 1: optimal k determination w/ fasthplus
##### adapting from louise H's script for dlPFC spatial: https://github.com/LieberInstitute/spatialDLPFC/blob/main/code/analysis/06_fasthplus/01_fasthplus.R

the 1310 svgs (umap plot 4) look cleanest so we'll use svgs as our feature set here.
```{r}
pcalist <- pcalist[[4]]
gc(full=T)
### we can do 5 at a time locally with a dataset this size, so:
kspan <- c(6:30)

i<-1
for (i in c(1:(length(kspan)/5))){
  ks <- kspan[c((5*(i-1)+1):(5*i))]
  multiklist <- list(pcalist,pcalist,pcalist,pcalist,pcalist)
  multiklist <- lapply(multiklist,FUN=function(x){
    reducedDimNames(x) <- "PCA"
    x})
  # hpb estimate. t = pre-bootstrap sample size, D = reduced dimensions matrix, L = cluster labels, r = number of bootstrap iterations
  
  # helper functions
  find_t <- function(L, proportion = 0.05) {
        initial_t <- floor(length(L) * proportion)
        smallest_cluster_size <- min(table(L))
        n_labels <- length(unique(L))
        ifelse(smallest_cluster_size > (initial_t / n_labels), initial_t, smallest_cluster_size * n_labels)
  }
  
  bpparam <- MulticoreParam(workers=8)
  register(bpparam)
  
  mk.glist <- bpmapply(function(X,Y){scran::buildSNNGraph(X,k=Y,use.dimred=reducedDimNames(X))}, BPPARAM = bpparam, X=multiklist, Y=ks)
  
  mk.g_walk <- bplapply(mk.glist,igraph::cluster_walktrap,BPPARAM = bpparam)
  rm(mk.glist)
  mk.walkclusts <- lapply(mk.g_walk,FUN=function(x){x$membership})
  rm(mk.g_walk)
  
  # assign as character, not factor -- otherwise we can't drop rows properly in fasth loop below
  multiklist <- bpmapply(X=mk.walkclusts,Y=multiklist,FUN=function(X,Y){
    colLabels(Y) <- factor(X)
    Y})
  # check if all this worked
  lapply(multiklist,FUN=function(x){table(colData(x)$label)})
  # ok
  
  multiklist <- mapply(X=multiklist,Y=ks,FUN=function(X,Y){
    colnames(colData(X))[which(colnames(colData(X))=="label")] <- paste0("label_",Y)
    X
  })
  
  fhres <- mapply(X=multiklist,Y=ks,function(X,Y){
    initial_t <- find_t(L = colData(X)[[paste0("label_",Y)]], proportion = 0.01)
    cluster_prop <- table(colData(X)[[paste0("label_",Y)]]) / ncol(X)
    bad_clusters <- which(cluster_prop < 0.01)
      if (length(bad_clusters) > 0) {
          message("For k: ", Y, " we are dropping small clusters: ", paste(names(bad_clusters), collapse = ", "))
          X <- X[, !(levels(colData(X)[[paste0("label_",Y)]]) %in% as.character(names(bad_clusters)))]
          updated_t <- find_t(colData(X)[[paste0("label_",Y)]], 0.01)
          message("initial t: ", initial_t, "; updated t: ", updated_t)
      }
      else{
        updated_t <- initial_t
      }
      
    set.seed(42)
    hpb(D = reducedDims(X)$PCA, L = colData(X)[[paste0("label_", Y)]], t = updated_t, r = 100)
  })
  
  if(i==1){fhrestab <- cbind(ks,unlist(fhres))}
  else{fhrestab <- rbind(fhrestab,cbind(ks,unlist(fhres)))}
  rm(curks,fhres,multiklist,mk.walkclusts)
  gc(full=T)
}
rm(i, kspan)

colnames(fhrestab) <- c("k","fasthplus")


# write.table(fhrestab,"analysis/ch13-clustering/fasthplus_results_1310svgs_k6-30.txt",sep='\t',quote=F,row.names=F,col.names = T)

fhrestab <- fread("analysis/ch13-clustering/fasthplus_results_1310svgs_k6-30.txt")

### https://github.com/LieberInstitute/spatialDLPFC/blob/main/code/analysis/06_fasthplus/02_segmented_inflection_point.R 
f2 <- lm(V2 ~ ks, data = fhrestab)

seg <- segmented(f2,
    seg.Z = ~ks,
    npsi = 1
)

seg2 <- segmented(f2,
    seg.Z = ~ks,
    npsi = 2
)

seg2$psi
seg$psi
## ^ convergence on k=~28
```

#### recluster with 1310, SVGS, 
```{r}

library(data.table)
require(colorout)
ColorOut()
options("styler.addins_style_transformer" = "biocthis::bioc_style()")
library(SpatialExperiment)
# library(ggspavis)
library(scater) # addPerCellQC
# library(nnSVG)
library(BiocParallel)
library(scran)
library(parallel)
# library(scCustomize)
library(fasthplus)
# library(STexampleData)
library(segmented)

```

load only objs we need for mem effish
```{r}
pcalist <- readRDS("processed/ch12/pcalist.RDS")
```

### Extending from ch13, part 1: optimal k determination w/ fasthplus
##### adapting from louise H's script for dlPFC spatial: https://github.com/LieberInstitute/spatialDLPFC/blob/main/code/analysis/06_fasthplus/01_fasthplus.R

the 1310 svgs (umap plot 4) look cleanest so we'll use svgs as our feature set here.
```{r}
pcalist <- readRDS("processed/ch12/pcalist.RDS")
svgpca <- pcalist[[4]]
reducedDimNames(svgpca) <- "PCA"

svg.glist28 <- scran::buildSNNGraph(svgpca,k=28,use.dimred="PCA")
colLabels(svgpca) <- factor(igraph::cluster_walktrap(svg.glist28)$membership)

### plot cluster assignments on UMAP, PCA, spatial
library(ggspavis)
svgpca <- runUMAP(svgpca)
reducedDims(svgpca)$UMAP <- as.data.frame(reducedDims(svgpca)$UMAP)
colnames(reducedDims(svgpca)$UMAP) <- c("UMAP1","UMAP2")
plotDimRed(svgpca,annotate="label",type = "PCA")
plotDimRed(svgpca,annotate="label",type = "UMAP")

plotSpots(svgpca,annotate = "label")

# saveRDS(svgpca,"processed/ch13/1310svgs_k28walktrap_clustered_spe.RDS")
```
