---
title: "ch14-markers aka cluster de"
author: "Bernie Mulvey"
date: "2023-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height = 10,fig.width = 7,include = FALSE)
knitr::opts_chunk$set(fig.width=7,fig.height=10)
#### sets tab autocompletion for directories to begin from the directory containing the .Rproj session, instead of the script's directory if different
knitr::opts_knit$set(root.dir = here::here())

library(ggplot2)
library(data.table)
library(Biostrings)
library(gridExtra)
require(colorout)
ColorOut()
options("styler.addins_style_transformer" = "biocthis::bioc_style()")

theme_set(theme_bw()+theme(axis.text.x = element_text(size = 14), axis.title.x = element_text(size = 16), axis.text.y = element_text(size = 14), axis.title.y = element_text(size =16), plot.title = element_text(size = 20,hjust=0.5), strip.text = element_text(size=18), legend.text = element_text(size=10), legend.title = element_text(size=11,hjust=0.5)))
```

we'll do this with two sets of data clustered using the top 1310 SVGs and k=28: walktrap clusters (unsupervised, spatially unaware) and bayesspace clusters with same.

```{r}
splc.wt <- readRDS("processed/ch13/1310svgs_k28walktrap_clustered_spe.RDS")
splc.bs <- readRDS("processed/ch13/1310svgs_k28_bayesspace.RDS")
```

the function findMarkers in scran is used with binomial tests (proportion of cells expressing per cluster; "This is a more stringent test than the default t-tests, and tends to select genes that are easier to interpret and validate experimentally." per osta). we also specify upregulated genes in order to identify positive markers per cluster.

pheatmap is used per OSTA; rownames are set to gene names for plotting with it.

```{r}
rownames(splc.bs) <- rowData(splc.bs)$gene_name
rownames(splc.wt) <- rowData(splc.wt)$gene_name

bs.markers <- findMarkers(splc.bs,test="binom",direction="up")
wt.markers <- findMarkers(splc.wt,test="binom",direction="up")

```

we are returned list objects, with each item named by the corresponding cluster.

lets examine a cluster of potential interest, bs.28.
we can pull out some top markers, of which there may be more than one per ranking in the column Top (see second line below: 2 gives us 11 genes, 5 gives 23. lets do 7 to be wild)
```{r}
bs28 <- bs.markers[[28]]
bs28.plotset <- bs28[bs28$Top<=7,]
bs28.plotset.fx <- getMarkerEffects(bs28.plotset)
pheatmap(bs28.plotset.fx)
```

### note that our cluster of interest, 28, is NOT in the heatmap.

we can also make violin plots per cluster per gene. notice this auto-defaults to logcounts.
```{r}
plotExpression(splc.bs,features = rownames(bs28.plotset),x="label")
dev.off()
```

i guess that's it.
