---
title: "chapters8-11"
author: "Bernie Mulvey"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
library(data.table)
library(SpatialExperiment)
library(ggspavis)
library(scater) # addPerCellQC
# library(STexampleData)
```

## look at components of an spe obj
```{r}
spe <- Visium_humanDLPFC()
# eh i kinda would prefer to look at something with multiple samples
# rm(spe)
# unloadNamespace(STexampleData)

library(WeberDivechaLCdata)
## oh lord have to update R and bioconductor and everything bbl
```

# OK let's see what a dataset (multiple samples) looks like.
```{r}
splc <- WeberDivechaLCdata_Visium()
dim(splc)
head(colData(splc))
tail(colData(splc))
# ah ok so there's just columns for each sample strung together.
```

### Ch9: QC / filtering
1. Is the tissue oriented correctly?
```{r}
ggspavis::plotSpots(splc)
# well, these are all distinct sections, so that's a good enough start

# we would do this to get rid of non-tissue spots, but that's already been done here. let's double check that there's no non-tissue spots remaining though
sum(colData(splc)$in_tissue!=1,na.rm=T)
# qed

dim(splc)
# and which is why we have a number of spots (20380) not divisible by 8

```

2. Mitochondrial reads--which were actually not used as a filter for the LC data so we can actually apply this step. Get a vector of features and whether or not they're mitochondrial
```{r}
# data.table derived style (won't work)
# mitogenes <- rowData(splc[rowData(splc)$gene_name %in% grep(rowData(splc)$gene_name,pattern="MT-|mt-",value=T)])$gene_name

# also WONT work-vector of gene names to include
mitogenesymbs <- grep(rowData(splc)$gene_name,pattern="MT-|mt-",value=T)


## these WILL work:
#1. full-length vector of T/F for matching MT-.*
mitogenes <- rowData(splc)$gene_name %in% grep(rowData(splc)$gene_name,pattern="MT-|mt-",value=T)

# 2. as coded in the tutorial
is_mito <- grepl("(^MT-)|(^mt-)", rowData(splc)$gene_name)
rowData(splc)$gene_name[is_mito]

```


#### now the next step, totalling reads per spot might need a vector of length nrow to work. let's find out.
```{r}
# working from the vector of mito gene symbols
splc2 <- addPerCellQC(splc,subsets=list(Mito=is_mito))
# as expected
splc3 <- addPerCellQC(splc,subsets=list(Mito=mitogenesymbs))
# THAT doesn't work though--expects indices
rm(splc2,mitogenesymbs,is_mito)


# these data now are in our coldata ('sample'/'cell' data)
colData(splc)$subsets_Mito_percent
sum(colData(splc)$subsets_Mito_percent>20)
# ouch
```

plot some qc stats
### coldata()$sum represents # of UMIs per spot
### standard threshold is 600 UMIs
```{r}
ggplot(as.data.frame(colData(splc)),aes(x=sum))+
  geom_histogram(binwidth = 600)

# again, ouch
```

### also look at sum PER # cells per spot. using ggspavis::plotQC. set threshold_y to the value we might plan to use--this function will draw on a horiz line to show what would be getting clipped off

### oh, LC data doesn't have this (but those values would usually be there from VistoSeg)
```{r}
# plotQC(splc,type = "scatter",metric_x = "cell_count",metric_y ="sum",threshold_y = 600)


```