---
title: "chapters8-11"
author: "Bernie Mulvey"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
library(data.table)
library(SpatialExperiment)
library(ggspavis)
library(scater) # addPerCellQC
# library(STexampleData)
```

## look at components of an spe obj
```{r}
spe <- Visium_humanDLPFC()
# eh i kinda would prefer to look at something with multiple samples
# rm(spe)
# unloadNamespace(STexampleData)

library(WeberDivechaLCdata)
## oh lord have to update R and bioconductor and everything bbl
```

# OK let's see what a dataset (multiple samples) looks like.
```{r}
splc <- WeberDivechaLCdata_Visium()
dim(splc)
head(colData(splc))
tail(colData(splc))
# ah ok so there's just columns for each sample strung together.
```

### Ch9: QC / filtering
1. Is the tissue oriented correctly?
```{r}
ggspavis::plotSpots(splc)
# well, these are all distinct sections, so that's a good enough start

# we would do this to get rid of non-tissue spots, but that's already been done here. let's double check that there's no non-tissue spots remaining though
sum(colData(splc)$in_tissue!=1,na.rm=T)
# qed

dim(splc)
# and which is why we have a number of spots (20380) not divisible by 8

```

2. Mitochondrial reads--which were actually not used as a filter for the LC data so we can actually apply this step. Get a vector of features and whether or not they're mitochondrial
```{r}
# data.table derived style (won't work)
mitogenes <- rowData(splc[rowData(splc)$gene_name %in% grep(rowData(splc)$gene_name,pattern="MT-|mt-",value=T)])$gene_name

# WILL work:
mitogenes <- rowData(splc)$gene_name %in% grep(rowData(splc)$gene_name,pattern="MT-|mt-",value=T)

# or as coded in the tutorial
is_mito <- grepl("(^MT-)|(^mt-)", rowData(splc)$gene_name)
rowData(splc)$gene_name[is_mito]

```


#### now the next step, totalling reads per spot might need a vector of length nrow to work. let's find out.
```{r}
# working from the vector of mito gene symbols
splc <- addPerCellQC(splc,subsets=list(Mito=is_mito))
# these data now are in our coldata ('sample'/'cell' data)


```